---
title: "Modified Score Test for Linear Mixed Models"
author: "Karl Oskar Ekvall"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Modified Score Test for Linear Mixed Models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Installation

The package can be installed from GitHub, using devtools.

```{r}
# devtools::install_github("koekvall/lmm-stest")
library(lmmstest)
```


# Setting

Suppose a vector of responses $Y\in \mathbb{R}^n$, matrix of predictors $X\in
\mathbb{R}^{n\times p}$, and design matrix $Z \in \mathbb{R}^{n\times q}$
satisfy, for some $\beta \in \mathbb{R}^p$,
$$
  Y = X\beta + ZU + E,
$$
where $U = [U_1, \dots, U_q]^T$ is a multivariate normal vector of random
effects with mean 0 and $E = [E_1, \dots, E_n]^T$ is multivariate normal with
mean zero and covariance matrix $\sigma^2 I_n$, independent of $U$. Assume also
the elements of $U$ are independent and that, for every $j \in \{1, \dots, q\}$,
there is a $k \in \{1, \dots, d\}$ such that $\mathrm{var}(U_j) = \lambda_k^2$,
$\lambda = [\lambda_1, \dots, \lambda_d]^T$. Thus, the elements of $\lambda$ are
standard deviations of the random effects. Several random effects can have the
same variance, in which case $d < q$, or each can have their own variance and,
then, $d = q$.

# Motivation

Many popular tests perform poorly when one or more of the scale parameters
$(\sigma, \lambda)$ are near or at zero. The modified score test statistic,
which replaces first derivatives of the log-likelihood by second when the former
are identically zero, is an exception. The modified score test statistic is
equal to the score test statistic standardized by expected Fisher information
when no scale parameters are identically zero, and can be obtained as the limit
of that test statistic otherwise.

# Example

Suppose that for $i = 1, \dots, N$ and $t = 1, \dots, T$,
$$
  Y_{i, t} = \beta_1 + \beta_2 X_{i, t} + U_{1, i} + U_{2, i} X_{i, t} + E_{i, t}.
$$
This model can be written in stacked form by letting $Y = [Y_{1, 1}, \dots, Y_{N, T}]^T$ and defining $X$ and $Z$
accordingly. In particular, $Z \in \mathbb{R}^{n \times 2N}$ with $n = NT$. The random effect $U_{1, i}$ is a random intercept shared by all observations with the same index $i$ and $U_{2, i}$ is a random slope. Let $\lambda_1$ and $\lambda_{2}$ be the standard deviations of $U_{1, i}$ and $U_{2, i}$, respectively. 

Consider the null hypothesis $(\sigma, \lambda_1, \lambda_2) = (1, 0.5, 0)$. The following code generates one realization of the model, assuming the null hypothesis, and tests this hypothesis using the proposed test.

```{r}
set.seed(seed)
  N <- 30
  N_T <- 10
  sigma_e <- 1
  sigma_u1 <- 0.5
  sigma_u2 <- 0
  Beta <- c(1, 1)
 
  # Generate data
  X <- matrix(runif(N * N_T, -1, 2) , nrow = N, ncol = N_T)
  U1 <- rnorm(N, sd = sigma_u1)
  U2 <- rnorm(N, sd = sigma_u2)
  E <- matrix(rnorm(N * N_T, sd = sigma_e), nrow = N, ncol = N_T)
  Y <- X + 1 + E
  for(ii in 1:n_i){
    Y[ii, ] <- Y[ii, ] + U1[ii] + U2[ii] * X[ii, ]
  }
  D <- data.frame(c(t(Y)), c(t(X)))
  D$time <- rep(1:n_t, n_i)
  D$obs <- as.factor(rep(1:n_i, each = n_t))
  names(D)[1:2] <- c("y", "x")
  
  # Fit model
  fit <- lme4::lmer(y ~ x + (1|obs) + (0 + x|obs), data = D, REML = FALSE)
  VC <- as.data.frame(lme4::VarCorr(fit))
  # Fit null model and get lrt stat
  Z_tall <- as.matrix(lme4::getME(fit, "Z"))
  X_tall <- lme4::getME(fit, "X")
  Lam_null <- diag(rep(c(sigma_u1, sigma_u2), each = n_i), n_i * 2)
  Sigma_null <- diag(sigma_e^2, n_i * n_t) + Z_tall %*% Lam_null^2 %*% t(Z_tall)
  R <- chol(Sigma_null)
  y_null <- qr.solve(t(R), D$y)
  X_null <- qr.solve(t(R), X_tall)
  beta_null <- qr.coef(qr(X_null), y_null)
  ll_null <- lmmstest::log_lik(y = D$y,
                               X = X_tall,
                               Z = Z_tall,
                               Beta = beta_null,
                               sigma = sigma_e,
                               lambda = c(sigma_u1, sigma_u2),
                               lam_idx = rep(1:2, each = n_i),
                               diffs = 0)[[1]]
  ll_alt <- lmmstest::log_lik(y = D$y,
                               X = X_tall,
                               Z = Z_tall,
                               Beta = unname(fixef(fit)),
                               sigma = VC[3, 5],
                               lambda = VC[1:2, 5],
                               lam_idx = rep(1:2, each = n_i),
                               diffs = 0)[[1]]
  lrt_stat <- 2 * (ll_alt - ll_null)
  lrt_p_val <- pchisq(lrt_stat, df = 3, lower.tail = FALSE)
  
  # Wald test
  finf <- lmmstest::fish_inf(y = D$y,
                      X = X_tall,
                      Z = Z_tall,
                      Beta = beta_null,
                      sigma = sigma_e,
                      lambda = c(sigma_u1, sigma_u2),
                      lam_idx = rep(1:2, each = n_i))
  e <- VC[c(3, 1, 2), 5] - c(sigma_e, sigma_u1, sigma_u2)
  wald_stat <- c(crossprod(e, finf[3:5, 3:5] %*% e))
  wald_p_val <- pchisq(wald_stat, df = 3, lower.tail = FALSE)
  
  # Score test for random effects
  c(lmmstest::score_test(
    y = D$y,
    X = X_tall,
    Z = Z_tall,
    Beta = unname(fixef(fit)),
    sigma = sigma_e,
    lambda = c(sigma_u1, sigma_u2),
    lam_idx = rep(1:2, each = n_i),
    test_idx = 2 + 1:3 # All variance parameters
  ),
```